Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2022-05-16 paper/ms.tex
FM and JQ edits

\begin{figure*}[h!]%
\centering
\includegraphics[width=18cm]{fig/fig1.pdf}
\caption{\textbf{Projected surface densities of simulated barred galaxies with
and without the interstellar medium.} The upper panels show an \Nbody{} only
simulation while the lower panels show a simulation which includes the SMUGGLE
model for the ISM. Columns show different points in time, separated by
$1\,\textrm{Gyr}$. We can see that in the \Nbody{} run, the bar grows in
length and strength. In the SMUGGLE run, the bar remains at approximately the
same length and strength over the course of the simulation. The \Nbody{} model
is identical to the GALAKOS model, discussed in the text.}\label{fig:overview}
\end{figure*}

@@

A surface density projection of our simulations is shown in
Fig.~\ref{fig:overview}. The upper three panels show the disk in the \Nbody{}
run while the lower three panels show the disk in the SMUGGLE run. Each column
shows the disk $\sim1\,\textrm{Gyr}$ apart in time. There is a large
qualitative difference in the evolution of the bar pattern between the two
runs. We see that in the \Nbody{} case, the bar lengthens in time and grows in
strength. In the SMUGGLE case, the bar retains a similar length and strength
between panels.

@@

The bottom panel of Fig.~\ref{fig:prop} shows the torque exerted on the bar by
different components. The solid lines indicate the torque on the bar by the
dark matter halo, whereas the dashed line indicates the torque on the bar by
the gas phase. In the \Nbody{} case, the halo exerts a steady negative torque on
the bar, with an average torque from $1$ to $4\,\textrm{Gyr}$ of $-58.0$ in
units of $10^{10}M_{\odot}\,(\textrm{km}/\textrm{s})^2$. The halo in the
SMUGGLE case exerts a similar negative torque on the bar in the first Gyr of
evolution, but after that the halo exerts a much smaller torque on the bar,
averaging only $-7.8$ in the same units and over the same interval. The gas in
the SMUGGLE case exerts a steady positive torque averaging $11.7$ in the same
units and over the same interval.

The fact that the dark matter halo in the SMUGGLE case exerts a smaller
positive torque on the bar can be understood in terms of the halo wake
mechanism. In the \Nbody{} case, halo material which is resonant with the bar
will form a wake which lags behind and exerts a negative torque on the bar,
which slows it down.\cite{1984MNRAS.209..729T, 1985MNRAS.213..451W,
1992ApJ...400...80H}\footnote{Since the bar is not a solid body, it is not
guaranteed that a negative torque will slow it down - e.g. a negative torque
could shred the bar, reducing its moment of inertia without changing its
pattern speed. However, the bar seems to empirically respond to a negative
torque induced by a halo wake by slowing down.} As the bar slows down, the
location of the resonances in phase space changes, allowing halo material
newly resonant with the bar to participate in forming the wake. However, the
gas is a reliable source of positive torque on the bar, speeding the bar up.
This stops the resonance location from changing such that the halo cannot
reinforce a wake, arresting the process by which the halo can slow the bar
down.

@@

We performed an experiment to test this hypothesis by forcing the stellar disk
in the SMUGGLE run to rotate at a constant angular rate and measuring the
torque on the bar by the gas phase at different rotation rates. The result of
this experiment is shown in Fig.~\ref{fig:equil}, which shows that a more
slowly rotating bar experiences a larger positive torque from the gas. We have
therefore shown evidence for an equilibrium mechanism which keeps the pattern
speed of the bar constant, resulting from the complex interplay between the
dark matter halo and the gas phase.

@@

Second, we performed a set of simulations in which we decreased the initial
gas mass of the disk. We essentially find for our disk, our proposed mechanism
operates as long as the gas contributes about $5\%$ of the total mass of the
disk. If the gas available is below this threshold, then the pattern speed
will decrease in a similar fashion to standard \Nbody{} simulations.
Importantly, arbitrarily increasing the gas fraction does not cause the
pattern speed to continuously increase. This is consistent with our proposed
equilibrium mechanism and with previous studies of barred galaxies which
varied the gas fraction.\cite{2010ApJ...719.1470V} The exact critical gas
fraction above which our mechanism operates is likely dependent on the
properties of the bar, disk, and dark matter halo -- further work is needed to
characterize how this gas fraction varies with such properties. Details on
this test in which we varied the gas mass of the disk are given in the Methods
section.

@@

\item [Acknowledgements] We would like to thank Greg~L. Bryan, Neal~J. Evans,
Drummond~B. Fielding, Keith Hawkins, Jason~A.~S. Hunt, Sarah~M.~R. Jeffreson,
Kathryn~V. Johnston, Peter~M.~W. Kalberla, Jürgen Kerp, Julio~F. Navarro,
Joshua~S. Speagle, Martin~D. Weinberg, and Yanfei Zou for helpful discussions.
A.B. would like to thank Todd Phillips for helpful discussions. Some
computations in this paper were run on the FASRC Cannon cluster supported by
the FAS Division of Science Research Computing Group at Harvard University.
Resources supporting this work were also provided by the NASA High-End
Computing (HEC) Program through the NASA Advanced Supercomputing (NAS)
Division at Ames Research Center. A.B. was supported by the Future
Investigators in NASA Earth and Space Science and Technology (FINESST) award
number 80NSSC20K1536 during the completion of this work. L.H. funding. E.D.
funding. F.M. funding. C.C. funding. J.Q. acknowledges support from NSF grant
AST-2008490. L.V.S. funding. P.T. acknowledges support from NSF grant
AST-1909933, AST-2008490, and NASA ATP Grant 80NSSC20K0502. M.V. acknowledges
support through NASA ATP 19-ATP19-0019, 19-ATP19-0020, 19-ATP19-0167, and NSF
grants AST-1814053, AST-1814259, AST-1909831, AST-2007355 and AST-2107724.

@@

\noindent
{\bf SMUGGLE Model}
\\
\noindent
We use the Stars and MUltiphase Gas in GaLaxiEs (SMUGGLE) model
\cite{2019MNRAS.489.4233M} implemented within the moving-mesh, finite-volume
hydrodynamics code AREPO \cite{2010MNRAS.401..791S}. The SMUGGLE model
includes self-gravity, hydrodynamics, radiative heating and cooling, star
formation, and stellar feedback. Explicit gas cooling and heating of the
multi-phase interstellar medium is implemented, covering temperature ranges
between $10$ and $10^8\,\textrm{K}$.

Star formation occurs in cells above a density threshold
($n_{\textrm{th}}=100\,\textrm{cm}^{-3}$) with a star-formation efficiency of
$\epsilon = 0.01$. Star formation converts gas cells into star particles which
represent single stellar populations. For each star particle, the deposition
of energy, momentum, and mass from stellar winds and supernovae is modeled.
Photo-ionization and radiation pressure are modeled using an approximate
treatment. A more detailed description of this model can be found in the
flagship SMUGGLE paper.\cite{2019MNRAS.489.4233M}

@@

\noindent
{\bf Varying Gas Fraction}
\\
\noindent
We also performed a test in which we varied the initial gas fraction of the
disk. In our fiducial run, we set the surface density of the gas disk from
$4\,\textrm{kpc}$ to $\sim9.3\,\textrm{kpc}$ to be $20\,\Msun/\textrm{kpc}^2$.
We also ran with surface densities of $15$, $10$, and
$5\,\Msun/\textrm{kpc}^2$. These correspond to initial gas fractions of
approximately $16\%$, $10\%$, $7\%$, and $4\%$. The pattern speed evolution is
shown in Extended Data Fig.~\ref{fig:fgas} We find that the bar in disks with
initial surface densities of $20$, $15$, and $10\,\Msun/\textrm{kpc}$ evolve
with a constant pattern speed while the bar in a disk with initial surface
density of $5\,\Msun/\textrm{kpc}^2$ slows down at a similar rate to the \Nbody{} case.

As a result, we conclude that for the disk, bar, and halo properties
considered in this work, a gas fraction of only approximately $5\%$ is
necessary in order for the proposed stabilizing mechanism to operate. Furthermore, we co

@@

\begin{figure}[t!]%
\centering
\includegraphics[width=9cm]{fig/fig-A2.pdf}
\caption{\textbf{Evolution of bar strength.} The bar strength is measured as
the maximum of the second Fourier component divided by the zeroth Fourier
component (a formula is given in the Methods section). We see that in the
\Nbody{} case (blue) the bar strength increases with time, consistent with
previous results showing that the bar strength increases as bars slow down. In
the SMUGGLE case (orange), we see that the bar strength remains roughly
constant, possibly slightly decreasing with time. This is also consistent with
the expected relation between pattern speed and strength since the bar in this
case is not slowing down.}
\label{fig:strength}
\end{figure}

@@

Computing the length of the bar and the torque on the bar by different
components requires us to decompose the disk into a component which is trapped
by the bar and a component which is untrapped. In order to do this, we follow
closely the technique developed in ref.\cite{2016MNRAS.463.1952P} We analyzed
the orbit of each star particle (meaning initial disk, bulge, and newly formed
stars) by extracting the $x$-$y$ positions of the apoapse of each in a frame
corotating with the bar (apoapses are defined as local maxima in $r$). For
each apoapse, we searched for the $19$ closest apoapses in time and applied a
$k$-means clustering algorithm on this set of $20$ points with $k=2$. We then
computed for each of the two clusters the average angle from the bar
$\left<\Delta \phi\right>_{0,1}$, the standard deviation in $R$ of the points
${\sigma_R}_{0,1}$, and the average radius of the cluster
$\left<R\right>_{0,1}$. At each apoapse, a particle was considered to be in
the bar if it met the following criteria:
\begin{equation}
\textrm{max}\left(\left<\Delta \phi\right>_{0,1}\right) < \pi / 8
\end{equation}
\begin{equation}
\frac{{\sigma_R}_0 + {\sigma_R}_1}{\left<R\right>_0 + \left<R\right>_1} < 0.22
\end{equation}
These criterion are slightly different and simplified from the ones used in
ref.\cite{2016MNRAS.463.1952P}, but we found to empirically work well at
decomposing the disk into a bar and disk component. In Extended Data
Fig.~\ref{fig:decomp}, we show an example of this decomposition. The
\textit{left} panel shows a surface density projection of the stellar disk and
bulge (including newly formed stars) from the SMUGGLE model after
$1\,\text{Gyr}$ of evolution in a frame such that the bar is aligned with the
$x$-axis. The \textit{middle} panel shows a projection of the subset of stars
that are identified as being trapped in the bar and the \textit{right} panel
shows a projection of the stars that are not identified as being trapped. The
fact that the \textit{right} panel is roughly axisymmetric indicates the bar
decomposition is performing adequately.

>>>>

\begin{figure*}[h!]%
\centering
\includegraphics[width=18cm]{fig/fig1.pdf}
\caption{\textbf{Projected surface densities of simulated barred galaxies with
and without the interstellar medium.} The upper panels show an \Nbody{} only
simulation while the lower panels show a simulation which includes the SMUGGLE
model for the ISM. Columns show different points in time, separated by about
$1\,\textrm{Gyr}$. We can see that in the \Nbody{} run, the bar grows in
length and strength. In the SMUGGLE run, the bar remains at approximately the
same length and strength over the course of the simulation. The \Nbody{} model
is identical to the GALAKOS model, discussed in the text.}\label{fig:overview}
\end{figure*}

@@

A face-on surface density projection of our simulations is shown in
Fig.~\ref{fig:overview}. The upper three panels show the disk in the \Nbody{}
run while the lower three panels show the disk in the SMUGGLE run. Each column
shows the disk $\sim1\,\textrm{Gyr}$ apart in time. There is a large
qualitative difference in the evolution of the bar pattern between the two
runs. We see that in the \Nbody{} case, the bar lengthens in time and grows in
strength. In the SMUGGLE case, the bar retains a similar length and strength
between panels.

@@

The bottom panel of Fig.~\ref{fig:prop} shows the torque exerted on the bar by
different components. The solid lines indicate the torque on the bar by the
dark matter halo, whereas the dashed line indicates the torque on the bar by
the gas phase. In the \Nbody{} case, the halo exerts a steady negative torque on
the bar, with an average torque from $1$ to $4\,\textrm{Gyr}$ of $-58.0$ in
units of $10^{10}M_{\odot}\,(\textrm{km}/\textrm{s})^2$. The halo in the
SMUGGLE case exerts a similar negative torque on the bar in the first Gyr of
evolution, but after that the halo exerts a much smaller torque on the bar,
averaging only $-7.8$ in the same units and over the same time interval. The gas in
the SMUGGLE case exerts a steady positive torque averaging $11.7$ over $1\,\textrm{Gyr}$ in the same
units.

The fact that the dark matter halo in the SMUGGLE case exerts a smaller
negative torque on the bar can be understood in terms of the halo wake
mechanism. In the \Nbody{} case, halo material which is resonant with the bar
will form a wake which lags behind and exerts a negative torque on the bar,
which slows it down.\cite{1984MNRAS.209..729T, 1985MNRAS.213..451W,
1992ApJ...400...80H}\footnote{Since the bar is not a solid body, it is not
guaranteed that a negative torque will slow it down - e.g. a negative torque
could shred the bar, reducing its moment of inertia without changing its
pattern speed. However, the bar seems to empirically respond to a negative
torque induced by a halo wake by slowing down.} As the bar slows down, the
location of the resonances in phase space changes, allowing halo material
newly resonant with the bar to participate in forming the wake. However, the
gas is a reliable source of positive torque on the bar, speeding the bar up.
This stops the resonance location from changing such that the halo cannot
reinforce a wake, arresting the process by which the halo can slow the bar
down.

@@

We performed an experiment to test this hypothesis by forcing the stellar disk
in the SMUGGLE run to rotate at a constant angular rate and measuring the
torque on the bar by the gas phase at different rotation rates. The result of
this experiment is illustrated in Fig.~\ref{fig:equil}, which shows that a more
slowly rotating bar experiences a larger positive torque from the gas. We have
therefore presented evidence for an equilibrium mechanism which keeps the pattern
speed of the bar constant, resulting from the complex interplay between the
dark matter halo and the gas phase.

@@

Second, we performed a set of simulations in which we decreased the initial
gas mass of the disk. We essentially find for our disk, our proposed mechanism
operates as long as the gas contributes about $5\%$ of the total mass of the
disk. If the gas available is below this threshold, then the pattern speed
will decrease in a similar fashion to standard \Nbody{} simulations.
Importantly, arbitrarily increasing the gas fraction does not cause the
pattern speed to continuously increase. This is consistent with our proposed
equilibrium mechanism and with previous studies of barred galaxies with varying gas fractions.\cite{2010ApJ...719.1470V} The exact critical gas
fraction above which our mechanism operates is likely dependent on the
properties of the bar, disk, and dark matter halo -- further work is needed to
characterize how this gas fraction varies with such properties. Details on
this test in which we varied the gas mass of the disk are given in the Methods
section.

@@

\item [Acknowledgements] We would like to thank Greg~L. Bryan, Neal~J. Evans,
Drummond~B. Fielding, Keith Hawkins, Jason~A.~S. Hunt, Sarah~M.~R. Jeffreson,
Kathryn~V. Johnston, Peter~M.~W. Kalberla, Jürgen Kerp, Julio~F. Navarro,
Joshua~S. Speagle, Martin~D. Weinberg, and Yanfei Zou for helpful discussions.
A.B. would like to thank Todd Phillips for helpful discussions. Some
computations in this paper were run on the FASRC Cannon cluster supported by
the FAS Division of Science Research Computing Group at Harvard University.
Resources supporting this work were also provided by the NASA High-End
Computing (HEC) Program through the NASA Advanced Supercomputing (NAS)
Division at Ames Research Center. A.B. was supported by the Future
Investigators in NASA Earth and Space Science and Technology (FINESST) award
number 80NSSC20K1536 during the completion of this work. L.H. funding. E.D.
funding. C.C. funding. J.Q. acknowledges support from NSF grant
AST-2008490. L.V.S. funding. P.T. acknowledges support from NSF grant
AST-1909933, AST-2008490, and NASA ATP Grant 80NSSC20K0502. M.V. acknowledges
support through NASA ATP 19-ATP19-0019, 19-ATP19-0020, 19-ATP19-0167, and NSF
grants AST-1814053, AST-1814259, AST-1909831, AST-2007355 and AST-2107724.

@@

\noindent
{\bf SMUGGLE Model}
\\
\noindent
We use the Stars and MUltiphase Gas in GaLaxiEs (SMUGGLE) model
\cite{2019MNRAS.489.4233M} implemented within the moving-mesh, finite-volume
hydrodynamics and gravity code AREPO \cite{2010MNRAS.401..791S}. The SMUGGLE model
additionally includes radiative heating and cooling, star
formation, and stellar feedback. Explicit gas cooling and heating of the
multi-phase interstellar medium is implemented, covering temperature ranges
between $10$ and $10^8\,\textrm{K}$.

Star formation occurs in cells above a density threshold
($n_{\textrm{th}}=100\,\textrm{cm}^{-3}$) with a star-formation efficiency of
$\epsilon = 0.01$. Star formation converts gas cells into star particles which
represent single stellar populations with an initial mass function according to ref.\cite{2001ApJ...554.1274C} For each star particle, the deposition
of energy, momentum, mass, and metals from stellar winds and supernovae is modeled.
Photo-ionization and radiation pressure are modeled using an approximate
treatment. A more detailed description of this model can be found in the
flagship SMUGGLE paper.\cite{2019MNRAS.489.4233M}

@@

\noindent
{\bf Varying Gas Fraction}
\\
\noindent
We also performed a test in which we varied the initial gas fraction of the
disk. In our fiducial run, we set the surface density of the gas disk from
$4\,\textrm{kpc}$ to $\sim9.3\,\textrm{kpc}$ to be $20\,\Msun/\textrm{pc}^2$.
We also ran with surface densities of $15$, $10$, and
$5\,\Msun/\textrm{pc}^2$. These correspond to initial gas fractions of
approximately $16\%$, $10\%$, $7\%$, and $4\%$. The pattern speed evolution is
shown in Extended Data Fig.~\ref{fig:fgas} We find that the bar in disks with
initial surface densities of $20$, $15$, and $10\,\Msun/\textrm{pc}^2$ evolve
with a constant pattern speed while the bar in a disk with initial surface
density of $5\,\Msun/\textrm{kpc}^2$ slows down at a similar rate to the \Nbody{} case.

As a result, we conclude that for the disk, bar, and halo properties
considered in this work, a gas fraction of only approximately $5\%$ is
necessary in order for the proposed stabilizing mechanism to operate.

@@

\begin{figure}[t!]%
\centering
\includegraphics[width=9cm]{fig/fig-A2.pdf}
\caption{\textbf{Evolution of bar strength.} The bar strength is measured as
the maximum of the second Fourier component divided by the zeroth Fourier
component. We see that in the
\Nbody{} case (blue) the bar strength increases with time, consistent with
previous results showing that the bar strength increases as bars slow down. In
the SMUGGLE case (orange), we see that the bar strength remains roughly
constant, possibly slightly decreasing with time. This is also consistent with
the expected relation between pattern speed and strength since the bar in this
case is not slowing down.}
\label{fig:strength}
\end{figure}

@@

Computing the length of the bar and the torque on the bar by different
components requires us to decompose the disk into a component which is trapped
by the bar and a component which is untrapped. In order to do this, we follow
closely the technique developed in ref.\cite{2016MNRAS.463.1952P} We analyzed
the orbit of each star particle (meaning initial disk, bulge, and newly formed
stars) by extracting the $x$-$y$ positions of the apoapse of each in a frame
corotating with the bar (apoapses are defined as local maxima in $r$). For
each apoapse, we searched for the $19$ closest apoapses in time and applied a
$k$-means clustering algorithm on this set of $20$ points with $k=2$. We then
computed for each of the two clusters the average angle from the bar
$\left<\Delta \phi\right>_{0,1}$, the standard deviation in $R$ of the points
${\sigma_R}_{0,1}$, and the average radius of the cluster
$\left<R\right>_{0,1}$. At each apoapse, a particle was considered to be in
the bar if it met the following criteria:
\begin{equation}
\textrm{max}\left(\left<\Delta \phi\right>_{0,1}\right) < \pi / 8
\end{equation}
\begin{equation}
\frac{{\sigma_R}_0 + {\sigma_R}_1}{\left<R\right>_0 + \left<R\right>_1} < 0.22
\end{equation}
These criterion are slightly different and simplified from the ones used in
ref.\cite{2016MNRAS.463.1952P}, but we found them to empirically work well at
decomposing the disk into a bar and disk component. In Extended Data
Fig.~\ref{fig:decomp}, we show an example of this decomposition. The
\textit{left} panel shows a surface density projection of the stellar disk and
bulge (including newly formed stars) from the SMUGGLE model after
$1\,\text{Gyr}$ of evolution in a frame such that the bar is aligned with the
$x$-axis. The \textit{middle} panel shows a projection of the subset of stars
that are identified as being trapped in the bar and the \textit{right} panel
shows a projection of the stars that are not identified as being trapped. The
fact that the \textit{right} panel is roughly axisymmetric indicates the bar
decomposition is performing adequately.
